---
title: "Week 3 - Project"
author: "Tim W"
date: "2/27/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Problem Description:

- Create a web page presentation using R Markdown that features a plot created with Plotly. Host your webpage on either GitHub Pages, RPubs, or NeoCities. 

- Your webpage must contain the date that you created the document, and it must contain a plot created with Plotly. 

-- We would love to see you show off your creativity!


## Review criteria

- The rubric contains the following two questions:

-- Does the web page feature a date and is this date less than two months before the date that you're grading this assignment?

-- Is the web page a presentation and does it feature an interactive plot that appears to have been created with Plotly?
    

## Approach:

Use earthquake data from the USGS site (https://earthquake.usgs.gov/earthquakes/feed/v1.0/csv.php), from Feb 27th, 2020, at 12:37pm EST.  Data source used is all earthquakes of magnitude 4.5 or greater, recorded over the past week.

Data source:  https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/4.5_week.csv




## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

