---
title: "Data Products - Project 2"
author: "Tim W"
date: "2/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Description:

1. Create a web page using R Markdown that features a map created with Leaflet.

2. Host your webpage on either GitHub Pages, RPubs, or NeoCities.

3. Your webpage must contain the date that you created the document, and it must contain a map created with Leaflet. We would love to see you show off your creativity! 

## Approach:

Use earthquake data from the USGS site (https://earthquake.usgs.gov/earthquakes/feed/v1.0/csv.php), from Feb 27th, 2020, at 12:37pm EST.  Data source used is all earthquakes of magnitude 4.5 or greater, recorded over the past week.

Data source:  https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/4.5_week.csv


```{r Load_Libraries}
# First things first, let's make sure those libraries are loaded.
# Of course, we need leaflet for map plotting.
library(leaflet)
# Let's also use htmltools to make formatting easier (lazy for the win, right?)
library(htmltools)
```

```{r Read_Data}
# Then, let's read in the CSV file pulled from the USGS site
QuakeData <- read.csv(file = "./4.5_week.csv", 
                    header = TRUE, 
                    sep = ",")

# This file is a comma separated value flat file, so separation character is a comma.
```

```{r Select_Columns}
# Now, let's grab the columns we care about from this spreadsheet.
# The columns of interest are:
# Place - a description of where the earthquake took place.
# Mag - the (Richter scale) magnitude of the earthquake.
# Lat - the latitude of the earthquake.
# long - the longitude of the earthquake.

map_data <- data.frame(place = QuakeData$place, 
                       mag = c(QuakeData$mag), 
                       lat = c(QuakeData$latitude), 
                       long = c(QuakeData$longitude))

# All data is read into a data frame called map data.
```


```{r Plot_Map}
# Finally, we have what we need to plot the map.
# We'll be using the pipe command to send data between the different functions of Leaflet
# Using variable called PlotMap to hold the map as we build it.

PlotMap <- map_data %>%
  leaflet() %>%
  addTiles() %>%
  addMarkers(popup=paste
             ("<br>Location: ", 
               htmlEscape(map_data$place), 
              "<br>Magnitude: ", 
               htmlEscape(map_data$mag))) 

# Our map will contain all earthquakes (of magnitude larger than 4.5) recorded
# over the past week.
# There are markers designating the name of where the earthquake took place
# as well as the recorded magnitude.

# Now, let's actually display the map.
PlotMap
```
                     
                     
