---
title: 'Reproducible Research: Peer Assessment 2'
output: html_document
keep_md: yes
---
## Introduction
Reproducible Research - Peer Assignment 2  
Author:  Tim Westran  
Purpose:  Answer questions about the NOAA Storm Database dataset  
Data set used:  https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2  
Github repo for the Course: [Reproducible Research](https://github.com/westrant/RepData_PeerAssessment2)  


## Synopsis
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. Various questions have been posed, and are answered below using the dataset.  Code is provided for the analysis performed.  

Questions answered:
1.  Across the United States, which types of events are most harmful with respect to population health? 
2.  Across the United States, which types of events have the greatest economic consequences?


## Data Processing
### Load libraries and set options
```{r includeLibraries}
## Ensure we have loaded the libraries we need.
library("data.table")
library("ggplot2")

## disable scientific notation
options(scipen = 999)
```

### Download data
Download the bz2 file and read it into a dataframe.
```{r DownloadAndReadData}
## Set the URL of the file to download
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

## Download the file to the working directory, save with filename:  repdata-data-StormData.csv.bz2
download.file(fileUrl, destfile = paste0("./", "repdata-data-StormData.csv.bz2"))

## Read in .bz2 file and store in a dataframe called stormData
stormData <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))

## Convert StormData dataframe to datatable called stormDataTable
stormDataTable <- as.data.table(stormData)
```

### Extract columns we care about.
We only need to keep some of the columns, so we remove everything except: EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP
```{r subsetData, results = "hide", cache = TRUE}
## Remove everything except EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP
columnsToDrop <- colnames(stormDataTable[, !c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")])

## Set all other columns to NULL.
stormDataTable[, c(columnsToDrop) := NULL]

## Grab all rows where a fatality or injury occurred, looking only at the columns we care about.
stormDataTable <- stormDataTable[(EVTYPE != "?" & (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP") ]
```


### convert dollars and exponents to a single value
To determine the impact of weather on the economy, we need to example property and crop damage figures.
Property damage is represented by PROPDMG, which contains the raw dollars, and PROPDMGEXP, which contains
the exponents.  We need to transform these two variables into a single value for each row.

Similarly, crop damage is represented by CROPDMG, which contains the raw dollars, and CROPDMGEXP, which 
contains the exponents.  We need to transform these two variables into a single value for each row.
```{r mergeValues, results = "hide", cache = TRUE}

## First, we modify all column names so they are all uppercase.
colNamesUpper <- c("PROPDMGEXP", "CROPDMGEXP")
stormDataTable[,  (colNamesUpper) := c(lapply(.SD, toupper)), .SDcols = colNamesUpper]

## Now, convert all property and crop damage exponent symbols to actual values.  The data set has different
## symbols mapped to different exponents.  
## First, property damage
propertyDmgConv <-  c("\"\"" = 1, "-" = 1, "+" = 1, "0" = 1, "1" = 10, "2" = 100,
                "H" = 100,  "3" = 1000,  "K" = 1000, "4" = 10000, "5" = 10^5, "6" = 10^6, 
                "M" = 10^6, "7" = 10^7, "8" = 10^8, "9" = 10^9, "B" = 10^9)

## Then crop damage exponent symbol mapping.
cropDmgConv <-  c("\"\"" = 1, "?" = 1, "0" = 1, "K" = 1000, "M" = 10^6, "B" = 10^9)

## Now, insert our converted exponents back into the data table.
## First, property damage.
stormDataTable[, PROPDMGEXP := propertyDmgConv[as.character(stormDataTable[,PROPDMGEXP])]]

## Setting all exponents with a value of NA to 1 (no exponent).
stormDataTable[is.na(PROPDMGEXP), PROPDMGEXP := 1 ]

## Then, property damage.
stormDataTable[, CROPDMGEXP := cropDmgConv[as.character(stormDataTable[,CROPDMGEXP])] ]

## Again, setting all exponents with a value Of NA to 1 (no exponent).
stormDataTable[is.na(CROPDMGEXP), CROPDMGEXP := 1 ]

## Add two new columns, propertyCost and cropCost
stormDataTable <- stormDataTable[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, propertyCost = PROPDMG * PROPDMGEXP, CROPDMG, CROPDMGEXP, cropCost = CROPDMG * CROPDMGEXP)]
```


### Calculate the total cost and fatalities / injuries
Here, we total up both the property and crop costs, then add all costs together to generate a total cost.
```{r totalCost, cache = TRUE}
## Create a new data table called totalCost that holds all summed cost figures.
totalCost <- stormDataTable[, .(propertyCost = sum(propertyCost), cropCost = sum(cropCost), Total_Cost = sum(propertyCost) + sum(cropCost)), by = .(EVTYPE)]

## Order from highest to lowest.
totalCost <- totalCost[order(-Total_Cost), ]

## Grab the top 10 costs and store them in the totalCost data table.
totalCost <- totalCost[1:10, ]

## Now, create a new data table called totalInjuriesAndFatailities that holds all of the summed injury
## and fatality statistics.
totalInjuriesAndFatailities <- stormDataTable[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), totals = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]

## Order from highest to lowest.
totalInjuriesAndFatailities <- totalInjuriesAndFatailities[order(-FATALITIES), ]

## Grab the top 10 injuries and fatalities, and store them in the data table.
totalInjuriesAndFatailities <- totalInjuriesAndFatailities[1:10, ]
```

## Results

### Health Inpact of Weather Events

Health impact of weather events - Weather events which are most harmful.
```{r healthImpactResults}
## First, create a new variable named harmfulEvents.
harmfulData <- melt(totalInjuriesAndFatailities, id.vars="EVTYPE", variable.name = "Impact")

## Create the chart to answer the question:  Which types of events are most harmful to population health?
healthChart <- ggplot(harmfulData, aes(x=reorder(EVTYPE, -value), y=value)) + 
    geom_bar(stat="identity", aes(fill=Impact), position="dodge") + 
    ylab("Count") + 
    xlab("Disaster") + 
    theme(axis.text.x = element_text(angle=45, hjust=1))+ 
    ggtitle("Most harmful weather events to population health") + 
    theme(plot.title = element_text(hjust = 1))

```

### Chart - Which weather events are most harmful?
```{r harmfulResultsDispHealth}
## Display the chart of harmful effects
healthChart
```

From the above chart, we can see that Tornados are the most harmful events, with 
`r format(totalInjuriesAndFatailities[1,2], big.mark = ",", scientific = FALSE)` fatalities, 
and `r format(totalInjuriesAndFatailities[1,3], big.mark = ",", scientific = FALSE)` injuries.

This is followed by excessive heat, with 
`r format(totalInjuriesAndFatailities[2,2], big.mark = ",", scientific = FALSE)` fatalities, 
and `r format(totalInjuriesAndFatailities[2,3], big.mark = ",", scientific = FALSE)` injuries.

With flash floods coming in 3rd, with 
`r format(totalInjuriesAndFatailities[3,2], big.mark = ",", scientific = FALSE)` fatalities, 
and `r format(totalInjuriesAndFatailities[3,3], big.mark = ",", scientific = FALSE)` injuries.

### Economic Inpact of Weather Events

Economic impact of weather events -  Weather events which have the greatest economic impact
```{r economicImpact}
## First, create a new variable named Cost.
economicImpact <- melt(totalCost, id.vars="EVTYPE", variable.name = "Cost")

## Create the chart to answer the question:  Which types of events are most harmful to population health?
economicImpactChart <- ggplot(economicImpact, aes(x=reorder(EVTYPE, -value), y=value)) +
    geom_bar(stat="identity", aes(fill=Cost), position="dodge") + 
    ylab("Cost") + 
    xlab("Disaster") + 
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("Most harmful weather events by economc impact") + 
    theme(plot.title = element_text(hjust = 1))
```

### Chart - Which weather events have the greatest economic impact?
```{r harmfulResultsDispEcon}
## Display the chart of harmful effects
economicImpactChart
```

From the above chart, we can see that floods are the most harmful events economically, 
with `r format(totalCost[1,2], big.mark = ",", scientific = FALSE)` in property damage, 
`r format(totalCost[1,3], big.mark = ",", scientific = FALSE)` in crop damage,
and `r format(totalCost[1,4], big.mark = ",", scientific = FALSE)` in total damage.

This is followed by hurricanes / typhoons, with 
`r format(totalCost[2,2], big.mark = ",", scientific = FALSE)` in property damage, 
`r format(totalCost[2,3], big.mark = ",", scientific = FALSE)` in crop damage,
and `r format(totalCost[2,4], big.mark = ",", scientific = FALSE)` in total damage.

With tornados coming in 3rd, with 
`r format(totalCost[3,2], big.mark = ",", scientific = FALSE)` in property damage, 
`r format(totalCost[3,3], big.mark = ",", scientific = FALSE)` in crop damage,
and `r format(totalCost[3,4], big.mark = ",", scientific = FALSE)` in total damage.

```{r datadisplay}
## Display top 10 most harmful events economically.
head(totalCost,10)

## Display top 10 most harmful events health-wise.
head(totalInjuriesAndFatailities,10)
```
